Reduced Nucleotide Diversity on a Plant Y Chromosome Following Recent Recombination Suppression

### 2016-08-16 manuscript.tex
minor edits

It is worth noting that our results only apply to the $Y_{1}$ chromosome, however, as our estimates of diversity were calculated using the set of sex-linked genes that were shared between the $XX/XY$ and the $XX/XY_{1}Y_{2}$ sex chromosome systems. Previous work investigating patterns of genetic degeneration found that both the $Y_{1}$ and the more recently evolved "neo" $Y_{2}$ chromosomes exhibit signs of genetic degeneration, including gene loss, loss of expression, and an accumulation of amino acid-changing mutations \citep{hough2014}, but we cannot say from the present study whether the neo-Y chromosome has also undergone a significant reduction in diversity. However, our results do indicate that the pattern of reduced diversity on the $Y_{1}$ chromosome occurs in each of the three \textit{R. hastatulus} sub-clades (\X Figure 2; Table 1; Figure S2), suggesting that this effect is not population-specific.

@@

Interestingly, our results indicate a significant reduction in X/A diversity in the derived SC and FL sub-clades of the North Carolina race ($X/A_{FL}=0.33$ and $X/A_{SC}=0.37$) compared to the Texas race ($X/A_{TX}=0.85$) (Figure 2). Although not expected, this reduction in diversity may be associated with the recent origin of the XY1Y2 sex chromosome system. The XY1Y2 sex chromosome system is thought to have originated through an X-autosome fusion (Smith 1964) involving the ancestral 3rd chromosome in the Texas race. Evidence supporting this autosomal origin was recently obtained by Grabowska-Joachimiak et al. (2014), who reported that the ancestral third chromosome in the Texas race carries the 5S rDNA locus, which is now found on both the neo-X and the Y2 sex chromosomes in the derived North Carolina race. If recent positive selection was responsible for driving the evolution of this fusion, then the formerly autosomal segment on the X chromosome in the XY1Y2 sub-clades (the neo-X) may have experienced a selective sweep, resulting in reduced X-linked diversity in the derived XY1Y2 sub-clades. It will be important for future work to investigate the factors driving the establishment of the X-autosome fusion in this species, and how they might impact patterns of X-linked neutral diversity.

In contrast to the situation on the X chromosome, our data indicate a strong and consistent diversity reduction on the Y chromosome, and we now test several possible models that might explain this diversity reduction.

\subsection*{Female biased sex ratios and variance in male fitness}

>>>>

It is worth noting that our results only apply to the $Y_{1}$ chromosome as estimates of diversity were calculated for sex-linked genes that were shared between the $XX/XY$ and the $XX/XY_{1}Y_{2}$ sex chromosome systems. Previous work found that both the $Y_{1}$ and the more recently evolved "neo" $Y_{2}$ chromosomes exhibited signs of genetic degeneration, including gene loss, loss of expression, and an accumulation of amino acid-changing mutations \citep{hough2014}, but we cannot say from the present study whether the neo-Y chromosome has also undergone a reduction in diversity. However, the extensive reduction in diversity estimated on the $Y_{1}$ chromosome occurs in each of the three \textit{R. hastatulus} sub-clades (\X Figure 2; Table 1; Figure S2), suggesting that this effect is not population-specific.

@@

Interestingly, our results also indicate a significant reduction in X/A diversity in the derived SC and FL sub-clades of the North Carolina race ($X/A_{FL}=0.33$ and $X/A_{SC}=0.37$) compared to the Texas race ($X/A_{TX}=0.85$) (Figure 2). Although not expected, this reduction in diversity may be associated with the recent origin of the $XY_{1}Y_{2}$ sex chromosome system, which is thought to have originated through an X-autosome fusion involving the ancestral 3rd chromosome in the Texas race \citep{smith1964evolving}. Evidence supporting this autosomal origin was recently obtained by \citep{grabowska2015}, who reported that the ancestral third chromosome in the Texas race carries the 5S rDNA locus, which is now found on both the neo-X and the $Y_{2}$ sex chromosomes in the derived North Carolina race. If recent positive selection was involved in driving the evolution of this X-A fusion, which theory suggests as likely \citep{charlesworth1980sex}, then the formerly autosomal segment on the X chromosome in the $XY_{1}Y_{2}$ sub-clades may have experienced a strong selective sweep, resulting in reduced X-linked diversity in the derived $XY_{1}Y_{2}$ sub-clades. It will be important for future work to investigate in more detail the factors driving the establishment of the X-autosome fusion in this species, and how they might impact patterns of X-linked neutral diversity.

In contrast to the situation on the X chromosome, our data indicate a strong and consistent diversity reduction on the Y chromosome, and we now consider several models that might explain reduction.

\subsection*{Female biased sex ratio and high variance in male fitness}
